\def\theTopic{Reading 7}


\begin{center}
{\bf {\large MIT -- the Male Idiot Theory}}
\end{center}


The usually  serious {\em British Medical Journal} enjoys a bit
of fun in each Christmas issue.  In December 2014 they
published a study of the MIT -- ``Males are Idiots Theory'' based on
data collected from the Darwin Awards.

``Winners of the Darwin Award must die in such an idiotic
manner that `their action ensures the long-term survival of the
species, by selectively allowing one less idiot to survive.'$^{20}$ The
Darwin Awards Committee attempts to make a clear distinction
between idiotic deaths and accidental deaths. For instance,
Darwin Awards are unlikely to be awarded to individuals who
shoot themselves in the head while demonstrating that a gun is
unloaded. This occurs too often and is classed as an accident.
In contrast, candidates shooting themselves in the head to
demonstrate that a gun is loaded may be eligible for a Darwin
Award--such as the man who shot himself in the head with a
`spy pen' weapon to show his friend that it was real.$^{18}$
To qualify, nominees must improve the gene pool by eliminating
themselves from the human race using astonishingly stupid
methods. Northcutt cites a number of worthy candidates.$^{12-21}$
These include the thief attempting to purloin a steel hawser from
a lift shaft, who unbolted the hawser while standing in the lift,
which then plummeted to the ground, killing its occupant; the
man stealing a ride home by hitching a shopping trolley to the
back of a train, only to be dragged two miles to his death before
the train was able to stop; and the terrorist who posted a letter
bomb with insufficient postage stamps and who, on its return,
unthinkingly opened his own letter.''\footnote{Lendrem, B. A. D.,
  Lendrem, D. W., Gray, A., \& Isaacs, J. D. (2014). The Darwin
  Awards: sex differences in idiotic behaviour. BMJ, 349, g7094.} 

The authors examined 20 years of data on the awards, removing awards
given to couples ``usually in compromising positions'' so that each
remaining winner was either male or female. Of the 318 remaining
awards, 282 were given to males and 36 were awarded to females.

They ask the question: ``If we look only at  people who do really
stupid things, what is the gender breakdown?''  or ``Are idiots more
than half male?''

\begin{center}
  {\large\bf Questions} 
\end{center}
\begin{enumerate}
\item What population is  represented by these winners of the
     Darwin Awards?
\begin{students}
    \vspace*{5cm}    
\end{students}
\begin{key}
  {\it One could answer that winners of the award are their own small
    population, and we have a census of all Darwin Award
    winners. However, there are other idiots who have not yet won this
    competition, but seem to be working toward proving themselves. I
    would say that the sample observed represents a population of people who
    take risks which should be avoided.} 
\end{key}
\item Rephrase the researchers' question in your own words.
\begin{students}
    \vspace{2cm}    
\end{students}
\begin{key}
  {\it What proportion of really idiotic humans are male?} 
\end{key}

\item What parameter would answer that question?
\begin{students}
    \vspace{1cm}    
\end{students}
\begin{key} 
  {\it $p$ = the proportion of all idiots who are male.}
\end{key}
\item What statistic gives us information about the parameter?  What
  is its value? (Use correct notation.)
\begin{students}
    \vspace*{2cm}    \\
\end{students}
\begin{key} 
   { $\widehat{p} = \frac{282}{318} = 0.887$ \it is the
    proportion of the sample which is male.}
\end{key}

\item Would the question be better answered with a confidence interval
  or a hypothesis test? Why?
\begin{students}
    \vspace*{2cm}    \\
\end{students}
\begin{key} 
   {\it  Could go either way. CI if they focused on estimating $p$,
     or a test if they focus on ``are idiots more than half male?''.}
\end{key}

\end{enumerate}
