\def\theTopic{Reading 15}


Watch the ``Correlation and Regression'' video at 
\url{https://camtasia.msu.montana.edu/Relay/Files/j63r458/Correlation_and_Linear_Regression/Correlation_and_Linear_Regression_-_20151020_090704_23.html }
\large
\begin{center}
  {\bf Important Points}
\end{center}

\begin{itemize}
\item What type of plot is used to show the relationship between two
  quantitative variables?\vspace{2cm}
\item What does correlation measure?\vspace{2cm}
\item What are possible values for correlation?  \vspace{2cm}
\item What symbol is used for population correlation?  for sample
  correlation? \vspace{2cm}  
\item What values for correlation indicate strong (weak) linear
  association?\vspace{2cm} 
\item  What symbols are used for the true parameters: intercept and
  slope in a regression model?\vspace{2cm}
\item  What symbols are used for the sample statistics: intercept and
  slope of the least squares line? \vspace{2cm}
\item  What is a residual? Which residuals are negative? positive?\vspace{2cm}
\item  We found the estimated least squares line to be
      $\widehat{Temp} = 37.7 +  0.23\times{Chirps}$ 
      \begin{itemize}
      \item Interpret the slope.\vspace{2cm}
      \item What is the predicted Temperature if we hear 100 chirps per
        minute? \vspace{2cm}
      \item What is the residual if $y = 58$ and $x =
        100$?\vspace{2cm}  
      \end{itemize}
\item Is it possible to have a negative slope and a positive
  correlation between two quantitative variables? Explain.\vspace*{\fill}
\end{itemize}
\normalsize