\def\theTopic{Effects of Concussion }
\def\dayNum{28 }

\begin{center}
{\bf {\large Concussions in the News}}\\
\end{center}

\begin{itemize}
\item 
Chris Borland, age 24, just retired from the 49ers after one year of
NFL play, because he fears his brain might be permanently injured by
concussions.
\item In 2013 the NFL agreed to pay \$765 million to fund exams,
  research and for compensation of former players.
\item High school coaches are now advised to make their
  players sit out if they sustain a blow to the head. 
\item The Congressional Brain Injury Task Force is looking into 
  the prevalence of brain trauma in troops deployed in recent wars and
  at how VA hospitals are treating the condition.
\end{itemize}

\begin{center}
  {\large\bf Applying What We've Learned}
\end{center}


 Some have argued that blows to the head -- for instance on the
    high school football field -- are getting too much attention.
    They point out that of many kids taking similar hits, only a few
    show lasting decrease in cognitive abilities, and suggest that the
    effects are as  much due to genetics as to the blow to the head.
    \begin{enumerate}
    \item Your group is asked to design a study to
      determine how large a risk the ``hits'' taken by high school
      athletes are to their future cognitive abilities. Write down
      a plan for your study.  Include:
      \begin{itemize}
      \item Who will your subjects be?  How will you find them? How
        many will you need? 
      \item What will you measure? (Brain scans like MRI?  Cognitive
        tests of memory and reasoning?  Some measure of emotional
        states like anger?) Include the timing of measurements.
      \item Will your study be observational or an experiment?
      \item Is your study ethical in its treatment of subjects?
      \end{itemize}
      %%  15 minutes?
   \newpage
      \vspace*{2.5in}

    \item Trade papers with another group and read their study over
      carefully. Address the bullet points above  (gently --
      remember someone is doing this to your proposal as well).
      Provide suggestions as to what might be improved.
      \begin{itemize}
      \item Subjects?
      \item Measurements?
      \item Observational? Experiment?
      \item Ethical issues?
      \end{itemize}
      \vspace{5in}
       %% 10 minutes?  

  
 % \item Why is it hard to figure out how large a role genetics play
 %      in the effects we observe?  \vspace{2in}

%   \item   Talavage, Thomas M., et al. ``Functionally-detected
%     cognitive impairment in high school football players without
%  clinically-diagnosed concussion.'' {\em Journal of Neurotrauma} {\bf
%    31}.4 (2014): 327-338.\\ 
% \url{http://online.liebertpub.com/doi/abs/10.1089/neu.2010.1512}
%   Obs study followed  11 HS players observed 
%   \begin{enumerate}
%   \item  no concussion \& no changes
%   \item  concussion \& changes
%   \item  no concussion \& changes
%   \end{enumerate}

%     What are the advantages of this study
%     design?  Disadvantages?  Does the design allow us to make causal
%     inferences?  To high school students?
    
%   \item 
% Brain Connect. 2015 Apr;5(2):91-101. doi: 10.1089/brain.2014.0279. Epub 2014 Oct 21.
% ``Alteration of default mode network in high school football athletes due to repetitive subconcussive mild traumatic brain injury: a resting-state functional magnetic resonance imaging study.''
% Abbas K1, Shenk TE, Poole VN, Breedlove EL, Leverenz LJ, Nauman EA,
% Talavage TM, Robinson ME.
% \url{http://www.ncbi.nlm.nih.gov/pubmed/25242171}

% \begin{quotation} 
% Abstract\\
% Long-term neurological damage as a result of head trauma while playing
% sports is a major concern for football athletes today. Repetitive
% concussions have been linked to many neurological disorders. Recently,
% it has been reported that repetitive subconcussive events can be a
% significant source of accrued damage. Since football athletes can
% experience hundreds of subconcussive hits during a single season, it
% is of utmost importance to understand their effect on brain health in
% the short and long term. In this study, resting-state functional
% magnetic resonance imaging (rs-fMRI) was used to study changes in the
% default mode network (DMN) after repetitive subconcussive mild
% traumatic brain injury. Twenty-two high school American football
% athletes, clinically asymptomatic, were scanned using the rs-fMRI for
% a single season. Baseline scans were acquired before the start of the
% season, and follow-up scans were obtained during and after the season
% to track the potential changes in the DMN as a result of experienced
% trauma. Ten noncollision-sport athletes were scanned over two sessions
% as controls. Overall, football athletes had significantly different
% functional connectivity measures than controls for most of the
% year. The presence of this deviation of football athletes from their
% healthy peers even before the start of the season suggests a
% neurological change that has accumulated over the years of playing the
% sport. Football athletes also demonstrate short-term changes relative
% to their own baseline at the start of the season. Football athletes
% exhibited hyperconnectivity in the DMN compared to controls for most
% of the sessions, which indicates that, despite the absence of symptoms
% typically associated with concussion, the repetitive trauma accrued
% produced long-term brain changes compared to their healthy peers.
%  \end{quotation}


\item   Read the abstract from this  article.

Petraglia AL, Plog BA, Dayawansa S, Chen M, Dashnaw ML, Czerniecka K,
Walker CT, Viterise T, Hyrien O, Iliff JJ, Deane R, Nedergaard M,
Huang JH. (2014). ``The spectrum of neurobehavioral consequences after repetitive mild traumatic brain injury: a novel mouse model of chronic traumatic encephalopathy.''
{\bf J Neurotrauma} Jul 1; {\bf 31}(13):1211-24. 
\url{http://www.ncbi.nlm.nih.gov/pubmed/24766454}
{\footnotesize
\begin{quotation}
Abstract\\
There has been an increased focus on the neurological consequences of
repetitive mild traumatic brain injury (TBI), particularly
neurodegenerative syndromes, such as chronic traumatic encephalopathy
(CTE); however, no animal model exists that captures the behavioral
spectrum of this phenomenon. We sought to develop an animal model of
CTE. Our novel model is a modification and fusion of two of the most
popular models of TBI and allows for controlled closed-head impacts to
unanesthetized mice. Two-hundred and eighty 12-week-old mice were
divided into control, single mild TBI (mTBI), and repetitive mTBI
groups. Repetitive mTBI mice received six concussive impacts daily for
7 days. Behavior was assessed at various time points. Neurological
Severity Score (NSS) was computed and vestibulomotor function tested
with the wire grip test (WGT). Cognitive function was assessed with
the Morris water maze (MWM), anxiety/risk-taking behavior with the
elevated plus maze, and depression-like behavior with the forced
swim/tail suspension tests. Sleep
electroencephalogram/electromyography studies were performed at 1
month. NSS was elevated, compared to controls, in both TBI groups and
improved over time. Repetitive mTBI mice demonstrated transient
vestibulomotor deficits on WGT. Repetitive mTBI mice also demonstrated
deficits in MWM testing. Both mTBI groups demonstrated increased
anxiety at 2 weeks, but repetitive mTBI mice developed increased
risk-taking behaviors at 1 month that persist at 6 months. Repetitive
mTBI mice exhibit depression-like behavior at 1 month. Both groups
demonstrate sleep disturbances. We describe the neurological consequenses
of repetitive mTBI in a novel mouse model, which resemble several of
the neuropsychiatric behaviors observed clinically in patients
sustaining repetitive mild head injury.
\end{quotation}
}
\begin{enumerate}
\item  What are the advantages of this study design?
\begin{students}
 \vfill
\end{students}

\begin{key}
  Replication, control, random assignment, very similar subjects.
\end{key}


\item Disadvantages?
\begin{students}
 \vfill
\end{students}

\begin{key}
 Does not extend to humans directly -- only if we assume some linkage.
\end{key}



\item Does the design allow us to make causal
    inferences?
\begin{students}
 \vspace{1cm}
\end{students}

\begin{key}
Yes
\end{key}



\item Inferences to high school students?

\begin{students}
\vspace*{.6cm}
\newpage
\end{students}

\begin{key}  No
\end{key}



\end{enumerate}

  \item And read this abstract:

Lin, Ramadan, Stern, Box, Nowinski, Ross, Mountford. (2015).
``Changes in the neurochemistry of athletes with repetitive brain trauma: preliminary results using localized correlated spectroscopy.''
{\bf Alzheimers Research \& Therapy}. 2015 Mar 15;7(1):13
\url{http://www.ncbi.nlm.nih.gov/pubmed/25780390}
{\footnotesize
\begin{quotation}
  Abstract\\
INTRODUCTION:\\
The goal was to identify which neurochemicals differ in professional
athletes with repetitive brain trauma (RBT) when compared to healthy
controls using a relatively new technology, in vivo Localized
COrrelated SpectroscopY (L-COSY).\\ 
METHODS:\\
To achieve this, L-COSY was used to examine five former professional
male athletes with 11 to 28 years of exposure to contact sports. Each
athlete who had had multiple symptomatic concussions and repetitive
sub concussive trauma during their career was assessed by an
experienced neuropsychologist. All athletes had clinical symptoms
including headaches, memory loss, confusion, impaired judgment,
impulse control problems, aggression, and depression. Five healthy
men, age and weight matched to the athlete cohort and with no history
of brain trauma, were recruited as controls. Data were collected from
the posterior cingulate gyrus using a 3 T clinical magnetic resonance
scanner equipped with a 32 channel head  coil.\\
RESULTS:\\
The variation of the method was calculated by repeated examination of
a healthy control and phantom and found to be 10\% and 5\%,
respectively, or less. The L-COSY measured large and statistically
significant differences (P $<=$0.05), between healthy controls and
those athletes with RBT. Men with RBT showed higher levels of
glutamine/glutamate (31\%), choline (65\%), fucosylated molecules
(60\%) and phenylalanine (46\%). The results were evaluated and the
sample size of five found to achieve a significance level P = 0.05.  Differences in N-acetyl aspartate and myo-inositol
between RBT and controls were small and were not statistically
significance.\\ 
CONCLUSIONS:\\
A study of a small cohort of professional athletes, with a history of
RBT and symptoms of chronic traumatic encephalopathy when compared
with healthy controls using 2D L-COSY, showed elevations in brain
glutamate/glutamine and choline as recorded previously for early
traumatic brain injury. For the first time increases in phenylalanine
and fucose are recorded in the brains of athletes with RBT. Larger
studies utilizing the L-COSY method may offer an in-life method of
diagnosis and personalized approach for monitoring the acute effects
of mild traumatic brain injury and the chronic effects of RBT. 
\end{quotation}
}
\begin{enumerate}
\item  What are the advantages of this study design?
\begin{students}
 \vfill
\end{students}

\begin{key}
 Ethically-- they didn't hurt anyone.
\end{key}



\item Disadvantages?
\begin{students}
 \vfill
\end{students}

\begin{key}
No random assignment.  Small samples.
\end{key}



\item Does the design allow us to make causal
    inferences?
\begin{students}
 \vspace{.6cm}
\end{students}

\begin{key} No
\end{key}



\item Inferences to high school students?
\begin{students}
\vspace*{.6cm}
\newpage
\end{students}

\begin{key} No
\end{key}

\end{enumerate}

\item Here's a quote from a news report in 2013:
{\footnotesize
  \begin{quotation}
    ``We need to figure out what's making some people more vulnerable
    than others,'' says Michelle Mielke, an Alzheimer's researcher at
    the Mayo Clinic who led the study. It was published online
    Thursday in the journal Neurology. 

``Just because you have a head trauma doesn't mean you're going to
develop memory problems or significant amyloid levels," Mielke told
Shots. And it doesn't mean you're going to get Alzheimer's. "But it
does suggest to us that there's a mechanism with head trauma that does
increase your risk.''

Mielke and her colleagues did PET scans on 589 people who are
participating in a long-term study of aging and memory. That's a lot
of people for a brain imaging study, which makes it more likely that
the findings are accurate. 
  \end{quotation}
}
\url{http://www.npr.org/blogs/health/2013/12/27/257552665/concussions-may-increase-}
\url{alzheimers-risk-but-only-for-some}

Explain the last sentence.  In what sense does a large sample increase accuracy?

\begin{students}
 \vspace{4cm}
\end{students}

\begin{key}
   Larger sample sizes give smaller variance estimates to $\xb$ and
   $|phat$.  That tends to shrink p--values and increase power, so we
   are less likely to make Type I or Type II errors. Larger samples do
   not decrease bias, so we'd still like to have random samples from
   the population.
\end{key}


\end{enumerate}


\begin{center}
  {\large\bf Take Home Message}
\end{center}

\begin{itemize}
\item  In reading a research article, carefully read what they say
  about selection of subjects (or units of study) so we know how far
  we can extend our inference.
\item Similarly, evaluate the assignment of any treatments (at random,
  we hope) so that we know if causal inference is appropriate.
\item In designing a study, we'd like to get a large sample, but
  expense might prevent that.
\item You might see names of tests we've not covered, but the idea of
  a p-value is the same for any test, and the null hypothesis is
  generally ``No change'', or ``No effect''.
\item What's not clear?  Write your questions here.
% P-values may be too small and confidence intervals may be too narrow
% if there is other variation present besides the ``sampling
% distribution''.  The methods you've learned in this class only cover
% the variation we get from randomly sampling subjects and randomly
% assigning treatments.
\end{itemize}


